# Graphs and plots
```{r setup7}
#| echo: false
#| message: false
#| warning: false

library(dplyr)
library(Hmisc)
library(ggplot2)
library(haven)
datadir<-"~/Data/bsa/"
bsa<-read_spss(paste0(datadir,"8849spss_V1/bsa2017_open_enviropol.sav"))
bsa$RAgeCat.f<-droplevels(as_factor(bsa$RAgeCat))
bsa$Rsex.f<-droplevels(as_factor(bsa$Rsex))
```
There are two common  ways to visualise data in R: either by using the straightforward but rather basic plotting commands from the Base package, or instead by delving into the more complex but much  nicer looking functionalities of the `ggplot` package. 

## Distributional graphs for continuous variables

Plots such as histograms or box plots are a convenient way to gain a quick overview of the distribution of a variable and are easy  to produce. Going back to the BSA data, we can plot the distribution of left-right political orientations scores with the `hist()` command. 

```{r 7.1}
#| fig-alt: "Unformatted histogram of left-right political orientations scores"
hist(bsa$leftrigh,freq=FALSE)
```

The histogram should appear in the ‘Plot’ tab on the right hand side of the R Studio window. It shows us that political orientations are slightly skewed towards the left. The `freq=FALSE` option requires the y-axis to be expressed in terms of proportions rather than number of observations.

Titles and labels can easily be added:
  
```{r 7.2}
#| fig-alt: "Histogram of left vs right political orientations, with title and y axis labels"
hist(bsa$leftrigh,
     freq=FALSE,
     main="Histogram of political orientations, 2017",
     ylab="Proportions",
     xlab="Left-right political orientations score")
```

Note that `main`, `ylab` and `xlab` can be used with any Base R plot commands. 

We can also produce a box and whisker plot of the same variable in order to get a better sense of the distribution of outliers:
```{r 7.3}
#| fig-alt: "Box and whisker plot of left vs right political orientations, with title and y axis labels"
boxplot(bsa$leftrigh,
        # main="Box and whisker plot of political orientations",
        ylab="Left-right political orientations score"
)
```

The generic `plot()` command produces scatterplots. Let's try it with our left-right political orientations score, in conjunction  with `libauth`, a libertarian vs authoritarian scale.
  
```{r 7.4.1}
#| fig-alt: "scatterplot of left-right political orientations vs authoritarianism score"

plot(bsa$leftrigh,bsa$libauth) ### Scatterplot of left-right political orientations score
```

The scatterplot shows us that there is little association between the two variables. However,  slightly fewer respondents simultaneously score  high on the 'authoritarianism' and 'right' scales, perhaps unsurprisingly.  

## Plotting categorical variables

The generic `plot()` function provides a quick way to produce bar plots of categorical data. For example, we can examine the distribution of political party affiliations (`Politics` variable). In order to do this, we convert it into a factor (i.e. categorical) variable as previously. Some preliminary abbreviation of the factor levels are also required in order for them to be displayed properly. 

```{r 7.5}
#| fig-alt: "Default frequency bar plot of political party affiliation using the plot command"

bsa$PartyId2.f<-droplevels( # Getting rid of unused factor levels for neater output
  as_factor(bsa$PartyId2)   # Converting haven labelled variable to factor
)
levels(bsa$PartyId2.f)<-c(
  "Con","Lab","Lib Dems","Other","None", "Greens" # Shorter level names
  ) 

plot(bsa$PartyId2.f)
```

More advanced plots require the barplot() function, which can be used in conjunction with table(). Whereas table() creates the data that will be plotted, barplot() does the actual plotting. For instance, we can produce the same bar plot, but this time with percentages, by creating a frequency table as we did above in Section 5.2, then plot it.

```{r 7.6}
#| fig-alt: "Neater frequency bar plot of political party affiliation using the barplot command" 

party.tab<-round(100*prop.table(
  table(bsa$PartyId2.f)
),
1)

party.tab

barplot(party.tab,
        main="Political party affiliation",
        ylab="Percent")
```

We can go further and create plots for two-way contingency tables of party affiliation by gender. This time we will do it in a single command:
  
```{r 7.7}
#| fig-alt: "Two-way frequency bar plot of political party affiliation by gender using the barplot command"

#t<-xtabs(~PartyId2.f+Rsex.f,bsa)       # First, let's get the contingency table
t<-xtabs(~Rsex.f+PartyId2.f,bsa)       # First, let's get the contingency table
barplot(
  round(100*
          prop.table(t,2),              ## Column % (here, gender)
    1),               ## Rounded to 1 decimal
  beside = T,     ## Side-by-side bars   
  main="Political party affiliation by gender",
  ylab="Percent")
```

## More advanced plots
Social science research often requires more advanced plots than just bar charts in order to conduct insightful analyses, such as for instance comparing the mean or median value of a continuous outcome across two or more categorical variables. The `ggplot` package  provides one of the most advanced set of tools for  data visualisation  currently available. A few examples are provided below.


#### Three way barplots using ggplot2 {-}

Say we would like to explore how differences in political party affiliations vary by gender and whether respondents have a degree-level education.

Let us first prepare the data: we need to create the table of result, the proportion of degree vs non degree holders by gender and political party. This is a three-way contingency table, that we can obtain with `ftable()` as shown in Section 5.2, combined with `prop.table()` for the computation of proportions and `round()`.

As they are more straightforward to handle in `ggplot`, we convert the table object created by `ftable` into a data frame. Although it is possible to specify titles and axis labels  in the plotting command, we will keep things simple and have them already in the data.  

Rather than using the full range of educational achievements recorded in `HEdQual3`, we would like instead to have a dichotomic variable measuring whether  respondents are degree holders or not. Adding it directly in the `ftable` command as a boolean expression return a dichotomic variable: "TRUE" for Degree educated respondents, and "FALSE" for everyone else. We just need to change the levels of this factor variable to make them more intelligible. Finally, we change the variable names in our data frame. 

```{r 7.9}
bsa$HEdQual3.f<-droplevels(as_factor(bsa$HEdQual3))
pa<-round(100*
            prop.table(
              ftable(bsa$PartyId2.f,bsa$Rsex.f,(bsa$HEdQual3.f=="Degree")
                     ),
              1),
          1)

pa<-data.frame(pa)
levels(pa$Var3)<-c("Below","Degree")
names(pa)<-c("Affiliation","Gender","Education","Percent")
pa
```
We are now ready to plot the data.  `ggplot()` functions usually work as a succession of layers or  options that are added to an initial plot specification. Each extra layer is added after a `+` sign. In the example below, we specify the data and the *aesthetic* (i.e. the basic parameters of the plot) with the first command: the x and y variables as well as the first grouping variable, in our case education). `geom_bar()` stipulates the bar plot, with the `ṕosition="dodge"` option for the bars to be located side by side (position="stack"would have them on top of each other). Finally, `facet_wrap()` splits the plot by gender.
```{r 7.10}
#| fig-alt: "Bar plot of political party affiliation by gender and qualification using ggplot"

ggplot(data=pa,aes(y=Percent,x=Affiliation,fill=Education))+
  geom_bar(position="dodge",stat="identity")+
  facet_wrap(~Gender)+
  theme_minimal()+                  ### Theme for visualisation
  scale_fill_manual(values=c("#702082", "#008755"))+ ### Custom colours (optional)
  theme(legend.position = "bottom")
```

#### Mosaic plots{-}
Mosaic plots provide a visualisation tool for two-way or more contingency tables. Table cells  are plotted as rectangles whose surface is proportional to their overall number of observations whereas their length represents their frequency relative to that of the second variable for each category of the first one - this is   equivalent to column percentages in a contingency table. In the example below, we are using the `mosaic()` function from the `vcd` package which provides both  descriptive and model-based plots. 

The basic parameters of a descriptive mosaic plot consist of:

- A R formula where the row and columns variables are specified;
- The highlighting variable, for the display of relative percentage;
- By default, the contrasting colours are shades of grey but can be customised, as can the direction of the bars.

Labelling options are a bit mode arcane than Base R plot functions. They need to be specified within a `labeling= labeling_border()` statement. See `help(labeling_border)` for more detail. We used `Varnames=F` to hide the  variable names from the plot, `rot_labels`, to specify that we did not want any rotation of the value labels, and `offset_labels` to prevent them to overlap with the rectangles.


```{r 7.8}
#| fig.height: 6
#| fig-align: center
#| fig-alt: "Mosaic plot of political party affiliation by gender "
library(vcd)                # Loading the vcd package

mosaic(~PartyId2.f+Rsex.f,                            
       data=bsa,                                      
       highlighting = "Rsex.f",                       
       labeling= labeling_border(varnames = F,            
                                 rot_labels = c(0,0,0,0),
                                 offset_labels = c(0, 0, 0, 1)
                                ), # labelling functions from mosaic. See 
                                   # help(labeling_border) for more detail
                    main="Mosaic plot of political party affiliation by gender" 
              )                          


```


\newpage